---
title: factor'em!
author: Olivier Bauthéac
date: ""
output: rmarkdown::html_vignette
bibliography: "../inst/REFERENCES.bib"
vignette: >
  %\VignetteIndexEntry{factor'em!}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style> body {text-align: justify} </style>
```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, eval = FALSE, comment = "#>")
```

factorem belongs to the [finRes](https://bautheac.github.io/finRes/) suite where it facilitates asset pricing research and factor investment backtesting. Install the development version with `devtools::install_github("bautheac/factorem")`.  

The package is organised around a workhorse function and a series of wrappers for asset pricing factors popular in the literature. The latter functions get raw financial data retrieved from Bloomberg using the [pullit](https://bautheac.github.io/pullit/) package and return S4 objects that carry data belonging to the corresponding factor including positions and return time series.  


## factorem

The factorem function is the workhorse function in [factorem](https://bautheac.github.io/factorem/). From raw financial data and a set of parameter specifications it contruct a complete backtest of the correpsonding asset pricing factor and returns the whole time series for the factor positions and return. For the record the S4 object returned by the function also contains the original raw financial and the set of parameters supplied as well as the original call to the function. The raw financial data should be supplied in a format similar to that of historical datasets returned returned by historical data query functions in pullit:

```{r `pullit equity BBG`, message = FALSE, warning = FALSE}
library(pullit); library(lubridate)

tickers_equity <- c("ADM US Equity", "CIVI US Equity", "GBX US Equity", "LIND US Equity", 
                    "SERV US Equity", "AE US Equity", "CLGX US Equity", "GDI US Equity", 
                    "LZB US Equity", "SGA US Equity", "AGCO US Equity", "CLR US Equity", 
                    "GHC US Equity", "MAN US Equity", "SITE US Equity", "AJRD US Equity", 
                    "COMM US Equity", "GME US Equity", "MEI US Equity", "SMP US Equity")
end <- Sys.Date(); start <- end - years(2L)

equity_data_market <- pull_equity_market(source = "Bloomberg", tickers_equity, start, end, verbose = FALSE)

get_data(equity_data_market)
```


```{r `pullit equity storethat`, echo = FALSE, message = FALSE, warning = FALSE, eval = F}
library(pullit); library(lubridate); storethat <- "../data-raw/storethat.sqlite"

tickers_equity <- c("ADM US Equity", "CIVI US Equity", "GBX US Equity", "LIND US Equity", 
                    "SERV US Equity", "AE US Equity", "CLGX US Equity", "GDI US Equity", 
                    "LZB US Equity", "SGA US Equity", "AGCO US Equity", "CLR US Equity", 
                    "GHC US Equity", "MAN US Equity", "SITE US Equity", "AJRD US Equity", 
                    "COMM US Equity", "GME US Equity", "MEI US Equity", "SMP US Equity")
end <- Sys.Date(); start <- end - years(2L)

equity_data_market <- pull_equity_market(source = "storethat", tickers = tickers_equity, 
                                         start = start, end = end, file = storethat, verbose = FALSE)

get_data(equity_data_market)
```

From there, constuct a bespoke asset pricing factor using the factorem function. I.e., an equity momentum factor:
```{r `factorem factor`, message = FALSE, warning = FALSE, eval = F}
library(factorem)

ranking_period = 1L
factor <- factorem(name = "factorem", data = pullit::get_data(equity_data_market),
                   ranking_period = ranking_period)
```


## wrappers

factorem provides wrapper methods for popular asset pricing factors in the literature. At the time of writing the factors covered in the package include, equity market and momentum as well as futures market, momentum, commercial hedging pressure (CHP), open interest and term structure factors. The author welcomes pull requests that would help expanding the current coverage.


### equity

#### market
The equity market factor encompasses the entire cross-section of a defined investment opportunity set with, at the time of writing, equal individual position weights. It came into sight in the 1960s for the central role it plays in the Capital Asset Pricing Model [@treynor1961market; @treynor1961toward; @sharpe1964capital; @sharpe1964capital; @mossin1966equilibrium; @lintner1975valuation] and remains the most popular factor to date in the literature when it comes to explaining the cross-section of equity returns:

```{r `equity market`, message = FALSE, warning = FALSE}
equity_market <- market_factor(data = equity_data_market)
```

#### momentum
The equity momentum factor is another popular factor in the asset pricing literature. After being first introduced in @carhart1997persistence it eventually drew the field leaders' attention in @fama2012size. The equity momentum sorts on prior returns over a defined ranking period:

```{r `equity momentum`, message = FALSE, warning = FALSE}
ranking_period = 1L
equity_momentum <- momentum_factor(data = equity_data_market, ranking_period = ranking_period)
```


### futures

```{r `pullit futures`, message = FALSE, warning = FALSE}
tickers_futures <- c("C A Comdty", "CCA Comdty", "CLA Comdty", "CTA Comdty", "FCA Comdty", 
                     "GCA Comdty", "HGA Comdty", "HOA Comdty", "KCA Comdty", "KWA Comdty", 
                     "LBA Comdty", "LCA Comdty", "LHA Comdty", "NGA Comdty", "O A Comdty", 
                     "PAA Comdty", "S A Comdty", "SIA Comdty", "W A Comdty", "XBA Comdty")

futures_data_TS <- pull_futures_market(source = "storethat", type = "term structure", tickers_futures, 
                                       start, end, file = storethat, verbose = FALSE)

get_data(futures_data_TS)
```


#### market
factorem provides futures equivalent for the factors above including a futures market factor that, at the time of writing, takes equally weighted positions on the nearby front contract of the commodity futures series provided:
```{r `futures market`, message = FALSE, warning = FALSE}
futures_market <- market_factor(data = futures_data_TS)
```


#### momentum
The momentum factor is also popular in the futures asset pricing literature in particular in the context of commodity markets [@miffre2007momentum]. As does its equity equivalent, it sorts on prior returns:
```{r `futures momentum`, message = FALSE, warning = FALSE}
ranking_period = 1L
futures_momentum <- momentum_factor(data = futures_data_TS, ranking_period = ranking_period)
```


#### commercial hedging pressure (CHP)
The futures commercial hedging pressure factor is based on the well-known hedging pressure-based theory [@anderson_hedger_1983; @chang_returns_1985; @cootner_returns_1960; @dusak_futures_1973; @hicks_value_1939; @hirshleifer_risk_1988; @hirshleifer_determinants_1989; @hirshleifer_hedging_1990; @kolb_is_1992; @keynes_treatise_1930] which postulates that futures prices for a given commodity are inversely related to the extent that commercial hedgers are short or long and the mimicking portfolio here aims at capturing the impact of hedging pressure as a systemic factor [@basu_capturing_2013].  
```{r `futures CHP`, message = FALSE, warning = FALSE}
futures_data_CFTC <- pull_futures_CFTC(source = "Bloomberg", tickers_futures, start, end, verbose = FALSE)

ranking_period = 1L
futures_CHP <- CHP_factor(price_data = futures_data_TS, CHP_data = futures_data_CFTC, 
                          ranking_period = ranking_period)
```


#### open interest
The open interest factor relates to futures market liquidity and is popular in the literature, in particular in the context of commodity markets research [@hong_what_2012]. It comes in two flavours in factorem, nearby and aggregate.

##### nearby
The nearby open interest factor is concerned with liquidity at the very front end of the term structure where it sorts on nearby front contract's open interest:
```{r `futures OI nearby`, message = FALSE, warning = FALSE}
ranking_period = 1L
futures_OI_nearby <- OI_nearby_factor(data = futures_data_TS, ranking_period = ranking_period)
```
##### aggregate
In contrast, the aggregate open interest factor [@basu_financiallization_2018] is concerned with liquidity over the whole futures term strucutre and sorts on open interest aggregated (summed up) over the futures term structure:
```{r `futures OI aggregate`, message = FALSE, warning = FALSE}
futures_data_agg <- pull_futures_market(source = "Bloomberg", type = "aggregate", tickers_futures, 
                                        start, end, verbose = FALSE)

ranking_period = 1L
futures_OI_aggregate <- OI_aggregate_factor(price_data = futures_data_market, 
                                            aggregate_data = futures_data_agg, 
                                            ranking_period = ranking_period)
```

#### term structure
The futures term structure factor [@szymanowska_anatomy_2014; @fuertes_commodity_2015] is concerned with the shape (steepness) of the futures term structure and sorts on roll yield:
```{r `futures TS`, message = FALSE, warning = FALSE}
ranking_period = 1L
futures_TS <- TS_factor(data = futures_data_market, ranking_period = ranking_period)
```


## accessors

Accessor methods allow access to the content of the factor output:
### name
```{r `factor name`, eval = F}
get_name(factor)
```
### positions
```{r `factor positions`, eval = F}
get_positions(factor)
```
### returns
```{r `factor returns`, eval = F}
get_returns(factor)
```
### data initially supplied
```{r `factor data`, eval = F}
get_data(factor)
```
### parameters
```{r `factor params`, eval = F}
get_parameters(factor)
```
### function call
```{r `factor call`, eval = F}
get_call(factor)
```


## summary
A summary method returns a performance summary of the corresponding facotr:
### function call
```{r `factor summary`, eval = F}
summary(factor)
```


## plotit
The [plotit](https://bautheac.github.io/plotit/) package, also part of the finRes suite, provides plot methods for the factor objects returned by the functions above:

### performance overview
```{r `plot performance`, fig.width = 7.5, fig.height = 5.5, fig.fullwidth = TRUE, eval = F, message = FALSE, warning = FALSE}
library(plotit)

plot(factor, type = "performance")
```

### positions overview
```{r `plot positions`, fig.width = 7.5, fig.height = 6.5, fig.fullwidth = TRUE, eval = F}
plot(factor, type = "positions")
```
  
  
  
  
## References



